\texttt{aeons}: approximating the end of nested sampling

### 2023-11-28 paper/paper.tex
name change

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics \citep{Trotta_2008}, materials science \citep{materials} and machine learning \citep{sparse_reconstruction}. The core algorithm is unique in its estimation of volumes by \textit{counting}, coupling together nested monte carlo integrals which makes high-dimensional integration feasible and robust. It also avoids problems that challenge traditional Bayesian samplers, such as posterior multi-modality and phase transitions.
\par
The order of magnitude run-time of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user. Currently, existing implementations of nested sampling (e.g. \citealt{multinest, polychord, dnest, dynesty, ultranest, nessai,proxnest}) either do not give an indication of remaining run-time, or only provide crude measures of progress that do not directly correspond to the the true endpoint.
\par
This paper sets out a principled manner of endpoint estimation for nested sampling at each intermediate stage (as shown in \cref{fig:polychord_output}), the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from. We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction in \cref{sec:anatomy}. \cref{sec:endpoint} then outlines the methodology we use, including discussion and comparisons to previous attempts. Finally, \cref{sec:results} presents the results and discussions for toy and cosmological chains, before we conclude.
\begin{figure}
%\begin{Verbatim}[frame=single, commandchars=\\\{\}]
%\textcolor{red}{Predicted endpoint: 25054 +/- 242}
%\textcolor{red}{Progress: [=================>########] 72%}
%___________________
%lives      |   500 |
%phantoms   | 24310 |
%posteriors | 18018 |
%equals     |   245 |
%-------------------
%ncluster   =  1/1
%ndead      =  18018
%nposterior =  18018
%nequals    =  249
%nlike      =  4159049
%<nlike>    =  491.04 (9.82 per slice)
%log(Z)     =  -12.55 +/- 0.27
%\end{Verbatim}
\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
\label{fig:polychord_output}
\end{figure}

@@

A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points , which is exactly needed to overcome the curse of dimensionality.

@@

A full picture of the distribution is visualised in grey in \cref{fig:logL_distribution}. 

@@

Suppose we are now at iteration $i$, with the latest dead point having likelihood $\log \Like_i$. As with the prior volumes, we shall attempt to quantify how much extra information about the likelihood is contained in the live points. For convenience, we will normalise the live point likelihoods as

@@

    \label{eq:y}

@@

so that $y=0$ corresponds to the latest dead point and $y=1$ to the maximum. At each step, $y$ increases by  

@@

@@

which is constant. The log-likelihood therefore proceeds towards the maximum geometrically, with a `time constant' of $dn_i/2$. We can again sum the harmonic series to find that the maximum live point will be at

@@

    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d},

@@

where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points. This is a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
The above also implies that the normalised distance between the highest and second highest live point is roughly 

@@

During the course of the run, the live points are always uniformly distributed over the constrained prior, and hence have probability distribution
    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
It is helpful at this stage to parameterise the live point log-likelihoods as
    \label{eq:y}
as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_i$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$. Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}\\ 
    [0<y_\mathrm{max}^\mathrm{live}<1],
    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
     \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.

@@

A more rigorous derivation (\cref{sec:logLmax}) shows that the maximum live point has mean and variance
\begin{equation}\label{eq:ylivemax}
	\lim_{d,\nlive\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log \nlive}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
Now let us take the dead point to be the termination point with likelihood $\log\Like_\mathrm{f}$ and prior volume $X_\mathrm{f}$, so that

@@

Putting this together, taking $\mathcal{L}_i$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$, and combining this with \eqref{eq:ylivemax} we find

@@

A plot of the log-likelihood distribution at the end of a run is shown in \cref{fig:logL_distribution}
    \includegraphics{figures/logL_distribution.pdf}
\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_i)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$. The distances are shown between these locations at the end of the run, the key takeaway being that in high dimensions the highest log-likelihood point of a nested sampling run is nowhere near the maximum in high dimensions.}
\label{fig:logL_distribution}

@@

\subsection{The termination prior volume}
Making the above discussion more precise, we wish to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}). The difficulty is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is far from the posterior bulk. 
\par
In order to get an idea of where the true posterior bulk sits, we need to predict what the posterior looks like past the highest live point. We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points. 
One would never use this predicted posterior to perform inference, since more accuracy can always be achieved by simply finishing the run. However, we will demonstrate it is sufficient for making a run-time prediction for $\log X_\mathrm{f}$. 
\par
Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
or equivalently
\begin{equation}\label{endpoint}
	\int_0^{X_\mathrm{f}} f(X, \phi)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \phi)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically. 
\par
Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule. For example in the constant $\nlive$ case at each iteration $\log X$ decreases by $1/n$, so the total number of iterations $N_\mathrm{f}$ would be
\begin{equation}
	N_\mathrm{f} = - n\log X_\mathrm{f} .
\end{equation}

\subsection{How to extrapolate?}
A key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian. Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function. The parameterisation of the Gaussian that we fit is the same as that given in \cref{sec:logL}, which we shall repeat here for clarity;
\begin{equation}\label{gaussian}
    f(X; \phi) = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
The extrapolation then proceeds thus:
\begin{enumerate}[leftmargin=*]
    \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior at the Bayesian temperature
    \item Take the live point profile and perform a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
    \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
\end{enumerate}
The advantage of fitting a Gaussian is that the procedure can be sped up analytically. Firstly, the least squares regression is trivial because analytic estimators exist; the cost function 
\begin{equation}\label{chi squared}
	C^2(\logLmax, \sigma) = \sum_i \left| \log \Like_i - f(X_i; \logLmax, \sigma) \right| ^2
\end{equation}
is minimised with respect to $(\logLmax, \sigma)$ when
\begin{equation}\label{eq:sigma}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i X_i^{2/d}\log \Like_i },
\end{equation}
and
\begin{equation}\label{eq:logLmax}
    \logLmax = \frac{1}{N} \sum_i \log \mathcal{L}_i + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d}.
\end{equation}
Secondly, the termination prior volume can also be obtained analytically. Rewriting \cref{endpoint} in terms of the Gaussian parameters gives
\begin{equation}
	\epsilon = \frac{\int_0^{X_\mathrm{f}} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X = \frac{d}{2} \cdot \left(\sqrt{2}\sigma\right)^d \cdot \gamma_k
\end{equation}
where $\gamma_k = \Gamma_{d/2}\left(X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{equation}
    \log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \Gamma^{-1}_{d/2} \left(\epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),\label{eq:xf}
\end{equation}
and $N_\mathrm{f}$ is of course just $-n$ multiplied by this. Intuitively, the above procedure can be thought of as inferring the number of constrained parameters, then extrapolating them up to find the point at which they will be fully constrained. 
\par
One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown this approach to be far less stable.

@@

\section{Results}\label{sec:results}
We now test the approach established in the previous section on a range of distributions. We begin by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
\subsection{Toy examples}
Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\cite{anesthetic}.
\subsubsection{Gaussians}
Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:gauss_predictions} as a benchmark for when fitting a Gaussian distribution is exact. In all endpoint prediction plots, the shading indicates the $1-2\sigma$ uncertainties. All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$. The correct endpoint is recovered to within standard error at all points except the very beginning, when the parameters have hardly been constrained. 
\par
We note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, which can be seen from the middle two plots.
\begin{figure*}
\begin{center}
    \includegraphics{figures/gauss_predictions.pdf}
\end{center}
\caption{Endpoint predictions for a spherical Gaussian run with $\nlive=500$ (except for the third plot from left). The correct endpoint is obtained for all but the earliest iterations, and the uncertainty is controlled by the number of live points, which can be seen from the two $d = 16$ plots.}
\label{fig:gauss_predictions}
\end{figure*}
We also observe the effect of elongating the Gaussian, using the same example as \cref{sec:elongated}. \cref{fig:elongated_logXfs} shows a step-like trend similar to the inferred dimensionalities, reflecting the fact that the full dimensionality is undetectable at lower compression factors. The endpoint for a likelihood whose remaining three directions are completely unconstrained coincides with our predictions at early iterations, showing that the two cases are indistinguishable. 
\begin{figure}
\begin{center}
    \includegraphics{figures/elongated_logXfs}
\end{center}
\caption{Endpoint prediction for an elongated Gaussian. At early stages, the full dimensionality is undetectable, and the endpoint is predicted to be the same as for a likelihood with three unconstrained directions. Only once the prior has been compressed enough to constrain the other three directions does the prediction converge to the true value.}
\label{fig:elongated_logXfs}
\end{figure}

\subsubsection{Cauchy}
One case that might be expected to cause problems is the pathological Cauchy distribution, which is far from a Gaussian. \cref{fig:cauchy_predictions} shows the predictions for a likelihood of the form
\begin{equation}
	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
\end{equation}
choosing $d = 10$ and allowing $\gamma$ to vary. The correct estimate is obtained to within standard error by about halfway, but before that is inaccurate. The key limitation is that the estimate is wrong early on, not because the compression is anisotropic, or that there is a phase transition; but rather as a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
\par
Nevertheless, the right order of magnitude is obtained at all times, so this remains sufficient for most use-cases. The Cauchy is also a pathological case, and the same problem does not in practice appear for more realistic cases, as we shall see next.
\begin{figure*}
\begin{center}
    \includegraphics{figures/cauchy_predictions.pdf}
\end{center}
\caption{Predictions for the Cauchy distribution for various widths $\gamma$ and $\nlive = 500$. The endpoint is underestimated for the first half of the run, but this is a limitation of the Gaussian approximation rather than a lack of information mid-run.}

@@

\section*{Code availability}
A package implementing the endpoint prediction mechanism is in development for easy plug-in to existing implementations of nested sampling. The latest updates can be found at \href{https://github.com/zixiao-h/aeons}{github.com/zixiao-h/aeons}.

>>>>

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics \citep{Trotta_2008}, materials science \citep{materials} and machine learning \citep{sparse_reconstruction}. The core algorithm is unique in its estimation of volumes by \textit{counting}, coupling together nested monte carlo integrals which makes high-dimensional integration feasible and robust. It also avoids problems that challenge traditional Bayesian samplers, such as posterior multi-modality and phase transitions.
\par
The order of magnitude run-time of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user. Currently, existing implementations of nested sampling (e.g. \citealt{multinest, polychord, dnest, dynesty, ultranest, nessai,proxnest}) either do not give an indication of remaining run-time, or only provide crude measures of progress that do not directly correspond to the the true endpoint.
\par
This paper sets out a principled manner of endpoint estimation for nested sampling at each intermediate stage (as shown in \cref{fig:polychord_output}), the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from. We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction in \cref{sec:anatomy}. \cref{sec:endpoint} then outlines the methodology we use, including discussion and comparisons to previous attempts. Finally, \cref{sec:results} presents the results and discussions for toy and cosmological chains, before we conclude.
\begin{figure}
\begin{Verbatim}[frame=single, commandchars=\\\{\}]
\textcolor{red}{Predicted endpoint: 25054 +/- 242}
\textcolor{red}{Progress: [=================>########] 72%}
___________________
lives      |   500 |
phantoms   | 24310 |
posteriors | 18018 |
equals     |   245 |
-------------------
ncluster   =  1/1
ndead      =  18018
nposterior =  18018
nequals    =  249
nlike      =  4159049
<nlike>    =  491.04 (9.82 per slice)
log(Z)     =  -12.55 +/- 0.27
\end{Verbatim}
\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
\label{fig:polychord_output}
\end{figure}

@@

A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points, which is exactly needed to overcome the curse of dimensionality.

@@

One can also derive the individual distributions for the live and dead points in log-likelihood. Note that this is merely the distribution of the \textit{values} of the points, without their nested sampling weights. The dead distribution is
    P(\log\Like) = \frac{\mathrm{d} \log X}{\mathrm{d} \log\Like} P(\log X) \propto \frac{dn}{2(\logLmax - \log\Like)},
since the log-densities of the dead point prior volumes are proportional to the live point number at any given iteration. Meanwhile, the live points are uniformly distributed, and so have the distribution 
    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}].
\cref{fig:logL_distribution} shows the distinction between the distributions of the live and dead point values, as well as the posterior. 

@@

How much further do the live points penetrate in log-likelihood? We seek the distribution for the maximum likelihood of the live points, $\log\Like_{(n_i)}^{\mathrm{live}}$, where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points.  For convenience, we will normalise the likelihoods as

@@

    \label{eq:normalised_likelihood}

@@

so that $y=0$ corresponds to the latest dead point at $\log\Like_i$ and $y=1$ to the maximum. The live point distribution simplifies to

@@

Using the result that the maximum of $n_i$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^{n_i})$, we obtain
    P(y_{(n_i)}^\mathrm{live}) = \frac{n_i d}{2}(1-y_{(n_i)}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_{(n_i)}^\mathrm{live})^{\frac{d}{2}}\right)^{n_i-1}\\ 
    [0<y_{(n_i)}^\mathrm{live}<1],
    y_{(n_i)}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n_i)}{\Gamma(1+\frac{2}{d}+n_i)} \\
     \pm \left( \frac{\Gamma(1+n_i)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n_i)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n_i)^2}{\Gamma(1+\frac{2}{d}+n_i)^2}\right)^{\frac{1}{2}},
    \lim_{d\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
    \lim_{d,n\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number. This is a very small fraction in high dimensions, showing that until the end the live points are far from the posterior bulk. 
Alternatively, the same result arrives intuitively from the fact that at each step, $y$ increases by  
\begin{equation}\label{eq:dy}

@@

so that by again summing the harmonic series, we get

@@

    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d}.

@@

\cref{eq:dy} also implies that the normalised distance between the highest and second highest live point is roughly 

@@

@@

Let us take the current iteration to be the termination point with likelihood $\log\Like_\mathrm{f}$ and prior volume $X_\mathrm{f}$, so that

@@

Putting this together, taking $\mathcal{L}_i$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$ and combining this with \eqref{eq:ymaxdn}, we find

@@

Making the termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence. However, nested sampling still shrinks around the maximum exponentially, so provided a good termination criterion is chosen, it will get there in reasonable time. What we are really interested in is the number of iterations required to get from the posterior bulk to the true maximum, which we will now calculate. At the posterior, we have roughly speaking
    (\log X, \log\Like) = (-\DKL, \logLmax - d/2).
The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum can then be found by inverting \cref{eq:gaussian_logL}:
    \log X_\delta = -\DKL - \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right),
which corresponds to an additional
    \Delta i = \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right)
iterations after the bulk is reached. One would therefore expect a general distribution to take $\mathcal{O}(d \log d)$ iterations to get from the usual nested sampling stopping point to within an e-fold of the maximum, which can serve as a sanity check for the maximisation procedure. 
A summary of the distances between the notable points at the end of a run is shown in \cref{fig:logL_distribution}.

\begin{figure*}
\begin{center}
    \includegraphics{figures/logL_distribution.pdf}
\end{center}
\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_i)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$. The distances are shown between these locations at the end of the run, the key takeaway being that in high dimensions the highest log-likelihood point of a nested sampling run is nowhere near the maximum in high dimensions.}
\label{fig:logL_distribution}
\end{figure*}

@@

\subsection{The termination prior volume}
Making the above discussion more precise, we wish to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}). The difficulty is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is far from the posterior bulk. 
\par
In order to get an idea of where the true posterior bulk sits, we need to predict what the posterior looks like past the highest live point. We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points. 
One would never use this predicted posterior to perform inference, since more accuracy can always be achieved by simply finishing the run. However, we will demonstrate it is sufficient for making a run-time prediction for $\log X_\mathrm{f}$. 
\par
Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
or equivalently
\begin{equation}\label{endpoint}
	\int_0^{X_\mathrm{f}} f(X, \phi)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \phi)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically. 
\par
Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule. For example, in the constant $\nlive$ case $\log X$ decreases by $1/n$ at each iteration, so the total number of iterations $N_\mathrm{f}$ would be
\begin{equation}
	N_\mathrm{f} = - n\log X_\mathrm{f} .
\end{equation}

\subsection{How to extrapolate?}
A key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian. Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function. The parameterisation of the Gaussian that we fit is the same as that given in \cref{sec:logL}, which we shall repeat here for clarity;
\begin{equation}\label{gaussian}
    f(X; \phi) = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
The extrapolation then proceeds thus:
\begin{enumerate}[leftmargin=*]
    \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior at the Bayesian temperature
    \item Take the live point profile and perform a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
    \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
\end{enumerate}
The advantage of fitting a Gaussian is that the procedure can be sped up analytically. Firstly, the least squares regression is trivial because analytic estimators exist; the cost function 
\begin{equation}\label{chi squared}
	C^2(\logLmax, \sigma) = \sum_i \left| \log \Like_i - f(X_i; \logLmax, \sigma) \right| ^2
\end{equation}
is minimised with respect to $(\logLmax, \sigma)$ when
\begin{equation}\label{eq:sigma}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i X_i^{2/d}\log \Like_i },
\end{equation}
and
\begin{equation}\label{eq:logLmax}
    \logLmax = \frac{1}{N} \sum_i \log \mathcal{L}_i + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d}.
\end{equation}
Secondly, the termination prior volume can also be obtained analytically. Rewriting \cref{endpoint} in terms of the Gaussian parameters gives
\begin{equation}
	\epsilon = \frac{\int_0^{X_\mathrm{f}} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X = \frac{d}{2} \cdot \left(\sqrt{2}\sigma\right)^d \cdot \gamma_k
\end{equation}
where $\gamma_k = \Gamma_{d/2}\left(X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{equation}
    \log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \Gamma^{-1}_{d/2} \left(\epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),\label{eq:xf}
\end{equation}
and $N_\mathrm{f}$ is of course just $-n$ multiplied by this. Intuitively, the above procedure can be thought of as inferring the number of constrained parameters, then extrapolating them up to find the point at which they will be fully constrained. 
\par
Uncertainties in the final estimate are obtained by drawing many samples from the distribution of $d_\mathrm{G}$ defined by the Bayesian temperature, and repeating step two for each. One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown this approach to be far less stable.

@@

\section{Results}\label{sec:results}
We now test the approach established in the previous section on a range of distributions. We begin by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
\subsection{Toy examples}
Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\citep{anesthetic}.
\subsubsection{Gaussians}
Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:gauss_predictions} as a benchmark for when fitting a Gaussian distribution is exact. In all endpoint prediction plots, the shading indicates the $1-2\sigma$ uncertainties. All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$. The correct endpoint is recovered to within standard error at all points except the very beginning, when the parameters have hardly been constrained. 
\par
We note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, which can be seen from the middle two plots.
\begin{figure*}
\begin{center}
    \includegraphics{figures/gauss_predictions.pdf}
\end{center}
\caption{Endpoint predictions for a spherical Gaussian run with $\nlive=500$ (except for the third plot from left). The correct endpoint is obtained for all but the earliest iterations, and the uncertainty is controlled by the number of live points, which can be seen from the two $d = 16$ plots.}
\label{fig:gauss_predictions}
\end{figure*}
We also observe the effect of elongating the Gaussian, using the same example as \cref{sec:elongated}. \cref{fig:elongated_logXfs} shows a step-like trend similar to the inferred dimensionalities, reflecting the fact that the full dimensionality is undetectable at lower compression factors. The endpoint for a likelihood whose remaining three directions are completely unconstrained coincides with our predictions at early iterations, showing that the two cases are indistinguishable. 
\begin{figure}
\begin{center}
    \includegraphics{figures/elongated_logXfs}
\end{center}
\caption{Endpoint prediction for an elongated Gaussian. At early stages, the full dimensionality is undetectable, and the endpoint is predicted to be the same as for a likelihood with three unconstrained directions. Only once the prior has been compressed enough to constrain the other three directions does the prediction converge to the true value.}
\label{fig:elongated_logXfs}
\end{figure}

\subsubsection{Cauchy}
One case that might be expected to cause problems is the pathological Cauchy distribution, which is far from a Gaussian. \cref{fig:cauchy_predictions} shows the predictions for a likelihood of the form
\begin{equation}
	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
\end{equation}
choosing $d = 10$ and allowing $\gamma$ to vary. The correct estimate is obtained to within standard error by about halfway, but before that is inaccurate. The key limitation is that the estimate is wrong early on, not because the compression is anisotropic, or because there is a phase transition; but rather as a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
\par
Nevertheless, the right order of magnitude is obtained at all times, so this remains sufficient for most use-cases. The Cauchy is also a pathological case, and the same problem does not in practice appear for more realistic cases, as we shall see next.
\begin{figure*}
\begin{center}
    \includegraphics{figures/cauchy_predictions.pdf}
\end{center}
\caption{Predictions for the Cauchy distribution for various widths $\gamma$ and $\nlive = 500$. The endpoint is underestimated for the first half of the run, but this is a limitation of the Gaussian approximation rather than a lack of information mid-run.}

@@

\section*{Code availability}
A package is in development to implement the endpoint prediction mechanism, for easy plug-in to existing implementations of nested sampling. The latest updates can be found at \href{https://github.com/zixiao-h/aeons}{github.com/zixiao-h/aeons}.
